### 2016-11-04 Paper_v8.tex
Added text to disucssion re Hong study

Systems biology and mathematical modeling are essential tools for understanding complex developmental programs like EMT \cite{Ahmed:2007mi}.
Previous computational models of TGF-$\beta$ induced differentiation focused on single biological factors or EMT in single cells.
For example, Chung \emph{et al.}, constructed a model of TGF-$\beta$ receptor activation and Smad signaling using ordinary differential equations and mass-action kinetics.
Their model suggested that a reduction of functional TGF-$\beta$ receptors in cancer cells may lead to an attenuated Smad2 signal \cite{Chung:2009jl}.
Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \cite{Vilar:2006gb}.
Several studies on ZEB signaling have modeled the role that micro-RNA can play in EMT, suggesting cell-fate determination is regulated by the miR-34/SNAIL and/or miR-200/ZEB mutual-inhibition feedback circuit \cite{Lu:2013aa,Zhang2014,Tian2013}.
Coarse grained modeling approaches have also been applied to EMT; Steinway \emph{et al.} used discrete dynamic modeling to study developmental EMT and known dysregulations in invasive hepatocellular carcinoma \cite{Steinway:2014aa}.
Although these models provided insight into the role of receptor dynamics, EMT induction involves many other components, including competing second messengers and interconnected transcriptional regulatory loops.
Integrating these additional scales of molecular signaling while maintaining the capacity for robust prediction requires a new and expanded computational and experimental strategy.
Data-driven systems approaches \cite{Cirit:2012kx} or logical model formulations \cite{Morris:2011ys} are emerging paradigms that constrain model complexity through the incorporation of training and validation data.
These are interesting techniques because the data informs model structure (which can be expanded as more data becomes available).
Alternatively, Bailey proposed more than a decade ago that a qualitative understanding of a complex biological system should not require complete definition of its structural and parametric content \cite{2001_bailey_NatBiotech}.
Shortly thereafter, Sethna and coworkers showed that complex model behavior is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as ``sloppiness'' \cite{Machta:2013by}.
Thus, reasonable model predictions are often possible with only limited parameter information.
Taking advantage of this property, we developed sloppy techniques for parameter identification using ensembles of deterministic models \cite{Song:2010fk}.
Furthermore, we proposed that the sloppy behavior of biological networks may also be seen as a source of cell-to-cell \cite{Lequieu:2011fj} or even patient-to-patient heterogeneity \cite{2010_luan_varner_MolBioSys}.
Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \cite{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
where a population of cells could be viewed as a dynamic ensemble of context-specific biochemical networks \cite{2012_creixell_linding_NatBiotech}.

@@

NF-$\kappa$B may also play an essential role of in the epithelial transformation.
NF-$\kappa$B may influence Snail expression through the AKT pathway and directly stabilize Snail activity \cite{Wu:2009cr}.  This is particularly important for integrating inflammation pathways, such as interleukin-6 (IL-6) and tumor necrosis factor-$\alpha$ (TNF-$\alpha$), which have been linked to EMT in pathological conditions \cite{Sullivan:2009bh}.
Other pathways such as Notch have also been shown to act synergistically with TGF-$\beta$ to express Slug in the developing embryo \cite{Niessen:2008nx}.
In addition, other important regulatory proteins such as ZEB1 should be added to the next generation model.
ZEB1 can activate the expression of mesenchymal genes such as collagens, smooth muscle actin and myosin, Vimentin while simultaneously contributing
to the repression of epithelial genes \cite{Liu:2008aa}.
Thus, ZEB1 can act in parallel to SNAIL and thus be an alternative to the canonical signaling paradigm presented here.
Lastly, while we have modeled classical protein signaling, we have not considered the role of regulatory RNAs on EMT.
There is growing evidence that microRNAs (miRNAs) play a strong role in EMT, where several miRNAs,
for example miR-21 and miR-31 are strongly associated with TGF-$\beta$ exposure \cite{Bullock:2012aa}, and have suggested that
the EMT process is driven by a miR-34/SNAIL and/or miR-200/ZEB feedback circuit \cite{Lu:2013aa,Zhang2014,Tian2013}.
Addressing missing structural components like these, could generate more insight into TGF-$\beta$ signaling and its role in phenotypic transformation.

@@

Our signaling network reconstruction was based on Medici et al. who identified the pathways though which MDCKII, DLD1 colon carcinoma,
and A375 melanoma cells transition towards a mesenchymal phenotype \cite{Medici:2008fk}.
Sequential activation of MAPK and Smad pathways were initiated upon addition of TGF$-\beta$1/2.
Briefly, TGF$-\beta$2 signals through the RAS-RAF-MEK-ERK pathway to up-regulate Snail and Slug expression \cite{Medici:2006qa}.
Snail, a known repressor of junctional proteins, inhibits the expression of E-cadherin \cite{Cano:2000kh}.
This initial repression of E-cadherin leads to a release of $\beta$-catenin from the cell membrane.
This release of $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$-\beta$3 expression \cite{Medici:2008fk}.
The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \cite{Medici:2008fk}.
GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway \cite{Zhou:2004kx,Larue:2005vn}.
Thereby, further $\beta$-catenin release also resulted from by TGF$-\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \cite{Derynck:2003fc}.
This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \cite{Massague:2005qc}.
In the model, receptors are simplified and represented as either bound or unbound complexes with their ligands.
Phosphorylated Smads 2/3 (pSmad2/3) form heterodimers and translocate to the nucleus. pSmads complexes up-regulate other transcription factors, such as LEF-1.
The pSmad2/4-LEF-1 complex has been shown to directly repress the E-cadherin gene \cite{Nawshad:2007kl}.
LEF-1 also binds with $\beta$-catenin to upregulate mesenchymal proteins such as fibronectin \cite{Medici2011}.
In the model, Smad signaling is consolidated into a single Smad species that can act in a co-dependent fashion with LEF1 to downregulate E-cadherin via a transfer function, eliminating the need for an explicity LEF-1, pSmad complex.
The EMT gene expression program was initiated by the binding of TGF$-\beta$ isoforms to TGF$-\beta$ surface receptors, starting the downstream signaling program.
Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \cite{Cano:2000kh}.
However, this transition is not solely driven by transcriptional events.
At the protein level, the repression of E-cadherin leads to a release of $\beta$-catenin from cell membrane.
Cystolic $\beta$-catenin then translocates to the nucleus and forms transcriptionally-active complexes with immunoglobulin transcription factor 2 (TCF-4)
to drive TGF$-\beta$3 expression \cite{Medici:2008fk}.
The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \cite{Medici:2008fk}.
GSK3 is known to act on $\beta$-catenin signaling through APC complex associated ubiquitin-proteasome pathway.
The APC complex is represented in our model and serves as a second reservoir of $\beta$-cateinin in untransformed cells whose sequestration is regulated by GSK3 \cite{Zhou:2004kx,Larue:2005vn,Medici:2008fk}.
Lastly, VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \cite{Suehiro:2014aa},
as supported by our VEGF-A experimental data (Fig. \ref{fg:VEGFA-Data}).

>>>>

Systems biology and mathematical modeling are essential tools for understanding complex developmental programs like EMT \cite{Ahmed:2007mi}.
Previous computational models of TGF-$\beta$ induced differentiation focused on single biological factors or EMT in single cells.
For example, Chung \emph{et al.}, constructed a model of TGF-$\beta$ receptor activation and Smad signaling using ordinary differential equations and mass-action kinetics.
Their model suggested that a reduction of functional TGF-$\beta$ receptors in cancer cells may lead to an attenuated Smad2 signal \cite{Chung:2009jl}.
Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \cite{Vilar:2006gb}.
Several studies on ZEB signaling have modeled the role that microRNA can play in EMT, suggesting cell-fate determination is regulated by the miR-34/SNAIL and/or miR-200/ZEB mutual-inhibition feedback circuits \cite{Lu:2013aa,Zhang:2014aa,Tian:2013aa}.
Coarse grained modeling approaches have also been applied to EMT; Steinway \emph{et al.} used discrete dynamic modeling to study developmental EMT and known dysregulations in invasive hepatocellular carcinoma \cite{Steinway:2014aa}.
Although these models provided insight into the role of receptor dynamics, EMT induction involves many other components, including competing second messengers and interconnected transcriptional regulatory loops.
Integrating these additional scales of molecular signaling while maintaining the capacity for robust prediction requires a new and expanded computational and experimental strategy.
Data-driven systems approaches \cite{Cirit:2012kx} or logical model formulations \cite{Morris:2011ys} are emerging paradigms that constrain model complexity through the incorporation of training and validation data.
These are interesting techniques because the data informs model structure (which can be expanded as more data becomes available).
Alternatively, Bailey proposed more than a decade ago that a qualitative understanding of a complex biological system should not require complete definition of its structural and parametric content \cite{2001_bailey_NatBiotech}.
Shortly thereafter, Sethna and coworkers showed that complex model behavior is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as ``sloppiness'' \cite{Machta:2013by}.
Thus, reasonable model predictions are often possible with only limited parameter information.
Taking advantage of this property, we developed sloppy techniques for parameter identification using ensembles of deterministic models \cite{Song:2010fk}.
Furthermore, we proposed that the sloppy behavior of biological networks may also be seen as a source of cell-to-cell \cite{Lequieu:2011fj} or even patient-to-patient heterogeneity \cite{2010_luan_varner_MolBioSys}.
Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \cite{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
where a population of cells could be viewed as a dynamic ensemble of context-specific biochemical networks \cite{2012_creixell_linding_NatBiotech}.

@@

NF-$\kappa$B may also play an essential role of in the epithelial transformation.
NF-$\kappa$B may influence Snail expression through the AKT pathway and directly stabilize Snail activity \cite{Wu:2009cr}.  This is particularly important for integrating inflammation pathways, such as interleukin-6 (IL-6) and tumor necrosis factor-$\alpha$ (TNF-$\alpha$), which have been linked to EMT in pathological conditions \cite{Sullivan:2009bh}.
Other pathways such as Notch have also been shown to act synergistically with TGF-$\beta$ to express Slug in the developing embryo \cite{Niessen:2008nx}.
In addition, other important regulatory proteins such as ZEB1 should be added to the next generation model.
ZEB1 can activate the expression of mesenchymal genes such as collagens, smooth muscle actin and myosin, Vimentin while simultaneously contributing
to the repression of epithelial genes \cite{Liu:2008aa}.
Thus, ZEB1 can act in parallel to SNAIL and thus be an alternative to the canonical signaling paradigm presented here.
Hong et al. developed an ODE model of ZEB1 signaling in MCF10A cells which also demonstrated intermediate EMT states \cite{Hong:2015aa}.
However, given that we do not have ZEB1 in our canonical architecture, this suggests multiple types of hybrid phenotypes could exist within the EMT program.
Lastly, while we have modeled classical protein signaling, we have not considered the role of regulatory RNAs on EMT.
There is growing evidence that microRNAs (miRNAs) play a strong role in EMT, where several miRNAs,
for example miR-21 and miR-31 are strongly associated with TGF-$\beta$ exposure \cite{Bullock:2012aa}, and have suggested that
the EMT process is driven by a miR-34/SNAIL and/or miR-200/ZEB feedback circuit \cite{Lu:2013aa,Zhang:2014aa,Tian:2013aa}.
Addressing missing structural components like these (and potentially others), could generate more insight into TGF-$\beta$ signaling and its role in phenotypic transformation.

@@

Our signaling network reconstruction was based on Medici et al. who identified the pathways though which MDCKII, DLD1 colon carcinoma,
and A375 melanoma cells transition towards a mesenchymal phenotype \cite{Medici:2008fk}.
Sequential activation of MAPK and Smad pathways were initiated upon addition of TGF$-\beta$1/2.
Briefly, TGF$-\beta$2 signals through the RAS-RAF-MEK-ERK pathway to up-regulate Snail and Slug expression \cite{Medici:2006qa}.
Snail, a known repressor of junctional proteins, inhibits the expression of E-cadherin \cite{Cano:2000kh}.
This initial repression of E-cadherin leads to a release of $\beta$-catenin from the cell membrane.
This release of $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$-\beta$3 expression \cite{Medici:2008fk}.
The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \cite{Medici:2008fk}.
GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway \cite{Zhou:2004kx,Larue:2005vn}.
Thereby, further $\beta$-catenin release also resulted from by TGF$-\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \cite{Derynck:2003fc}.
This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \cite{Massague:2005qc}.
In the model, receptors are simplified and represented as either bound or unbound complexes with their ligands.
Phosphorylated Smads 2/3 (pSmad2/3) form heterodimers and translocate to the nucleus. pSmads complexes up-regulate other transcription factors, such as LEF-1.
The pSmad2/4-LEF-1 complex has been shown to directly repress the E-cadherin gene \cite{Nawshad:2007kl}.
LEF-1 also binds with $\beta$-catenin to upregulate mesenchymal proteins such as fibronectin \cite{Medici2011}.
In the model, Smad signaling is consolidated into a single Smad species that can act in a co-dependent fashion with LEF1 to downregulate E-cadherin via a transfer function, eliminating the need for an explicity LEF-1, pSmad complex.
The EMT gene expression program was initiated by the binding of TGF$-\beta$ isoforms to TGF$-\beta$ surface receptors, starting the downstream signaling program.
Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \cite{Cano:2000kh}.
However, this transition is not solely driven by transcriptional events.
At the protein level, the repression of E-cadherin leads to a release of $\beta$-catenin from cell membrane.
Cytosolic $\beta$-catenin then translocates to the nucleus and forms transcriptionally-active complexes with immunoglobulin transcription factor 2 (TCF-4)
to drive TGF$-\beta$3 expression \cite{Medici:2008fk}.
The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \cite{Medici:2008fk}.
GSK3 is known to act on $\beta$-catenin signaling through APC complex associated ubiquitin-proteasome pathway.
The APC complex is represented in our model and serves as a second reservoir of $\beta$-catenin in untransformed cells whose sequestration is regulated by GSK3 \cite{Zhou:2004kx,Larue:2005vn,Medici:2008fk}.
Lastly, VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \cite{Suehiro:2014aa},
as supported by our VEGF-A experimental data (Fig. \ref{fg:VEGFA-Data}).
